---
title: "Saving and loading data"
---

At some point, we either want to load data into R or save a data frame that we have created or modified. Arguably, the best format to store data in tabular form (rows representing observations and columns representing variables) is the *CSV* (comma-separated variables) format.

CSV files are ideal for **Open Data** as most software can read them, including generic text editors.

In order to load and save data in *CSV* format, we can use two functions:

- *write.csv* will save tabular data (usually a data frame, but can also be a matrix) as a *CSV* file
- *read.csv* will load data from a *CSV* file


## Saving data frames

If we save a data frame using the function *write.csv*, we should specify three function arguments:

- *x*: the name of the object we want to save (a data frame, but works with matrices, too)
- *file*: the **directory** where we want to save the file **plus the file name** as a character string
- *row.names*: whether we want R to assign row names (default is TRUE, but we want to turn that to FALSE)

If we do not specify a directory, R will save the data frame in the **current working directory**.

We can ask R to tell us the current working directory using the function *getwd*. This function is special in that it does not have any function arguments. Here is what the syntax looks like:

```{r}
#| eval: false
#| echo: true
# Obtain the current working directory
getwd()
```

The output in the console will look slightly different for each person due to differences in how their folders are named:

:::{.alert .alert-warning}
```{r}
#| eval: true
#| echo: false
# Obtain the current working directory
print("C:/Users/Thomas/Documents/R stuff")
```
:::


Here is how the code for saving a data frame could look like.

```{r}
#| eval: false
#| echo: true
# save the object called my_data_frame
write.csv(
  x = my_data_frame,
  file = "C:/Users/Thomas/Documents/data.csv",
  row.names = FALSE
)
```

Provided the directory "C:/Users/Thomas/Documents" already exists on the computer, there will now be a file named "data.csv" in that directory.


## Loading data frames

We can load a data frame (or matrix) saved in *CSV* format using the function *read.csv*.

We usually need to specify only two function arguments:

- file: the **directory and file name** of the *CSV* file
- header: a Boolean variable indicating whether the first row of file contains the variable names (default is TRUE)

:::{.alert .alert-danger}
**Important**: If we simply call the function, R will display the data frame we loaded in the console. If we want it to appear in the **Environment** we need to define it as a new object.
::: 

Here is what the code would look like if we were to load the data frame saved in the previous example.

```{r}
#| eval: false
#| echo: true
# load data
my_data_frame = read.csv(
  file = "C:/Users/Thomas/Documents/data.csv",
  header = T)
```

Assuming that the requested file exists in the specified directory, the Environment will now show the object called "my_data_frame" in the Environment.


